\subsection{Gr{\"o}bner reduction}

Z3 uses a best effort Gr\"obner basis reduction to find inconsistencies, cheaply, 
and propagate consequences. While Gr{\"o}bner basis heuristics are not new to Z3,
they have evolved and to our knowledge the integration is unique among SMT solvers.
Recall that reduced Gr{\"o}bner basis for a set of polynomial equations $p_1 = 0, \ldots, p_k = 0$
is a set $q_1 = 0, \ldots, q_m = 0$, such that every $p_i$ is a linear sum of $q_j$'s,
and the leading monomials of every pair $q_i, q_j$, $i \neq j$, have no common factors.
Since Z3 uses completion as a heuristic to make partial inferences, it does not seek to compute
a basis.
The Gr\"obner module performs a set of partial completion steps, preferring
to eliminate variables that can be isolated, and expanding a bounded number of super-position
steps (reductions by S-polynomials).

Z3 first adds equations $m = x_1\ldots x_k$ for monomial definitions that are violated.
It then traverses the transitive cone of influence of Simplex rows that contain one of the added variables
from monomial definitions. It only considers rows where the basic variable is bounded.
Rows where the basic variable is unbounded are skipped
because the basic variable can be solved for over the reals.
Fixed variables are replaced by constants, and the bounds constraints that fixes the variables
are recorded as dependencies with the added equation. Thus, the equations handled by the
Gr{\"o}bner basis reduction are of the form $\langle p_i: xy + 3z + 3 = 0, d_i: \{3 \leq u \leq 3\} \rangle$, where $p_i$
is a polynomial and $d_i$ is a set of dependencies corresponding to fixed variables that were replaced
by constants in $p_i$. In the example, we replaced $u$ by $3$ and the definition $\langle m = xy, \emptyset\rangle$ resolved $m$ by $xy$.
Dependencies are accumulated when two polynomials are resolved to infer a new derived equality.
Generally, when $\langle xy + p_1 = 0, d_1\rangle, \langle xz + p_2 = 0, d_2 \rangle$ are two polynomial equations,
then $\langle zp_1 - yp_2 = 0, d_1 \cup d_2\rangle$ can be derived accumulating the premises $d_1, d_2$.

Finally, equations are pre-solved if they are linear and can be split
into two groups, one containing a single variable that has a
lower (upper) bound, the other with more than two variables
with upper (lower) bounds. This avoids losing bounds information
during completion.

After (partial) completion, the derived equations are post-processed:
\begin{description}

\item[Constant propagation]
For equalities of the form $x = 0$ or $ax + b = 0$. If the current assignment to $x$ does not
satisfy the equation, then the equality is propagated as a lemma.

\item[Linear propagation]
As a generalization of constant propagation, if the completion contains linear equations that
evaluate to false under the current assignment, then these linear equations are added to the
Simplex Tableau. Example~\ref{ex:yoav} illustrates a use where this propagation is useful.

\item[Factorization] Identify factors of the
form $x y p \simeq 0$ where $x, y$ are variables an $p$ is linear. We infer 
the clause $x y p \simeq 0 \implies x \simeq 0 \lor y \simeq 0 \lor p \simeq 0$.

\end{description}

\begin{example}[Combining Gr{\"o}bner completion and Linear Solving]
\label{ex:yoav}
We include an example obtained from Yoav Rodeh at Certora.
The instance was not solvable prior to adding simplex propagation.
To solve it, Certora relied on treating multiplication as an uninterpreted function and
including selected axioms for modular arithmetic and multiplication that were instantiated by E-matching.
The distilled example is:

\[
   L \leq x \cdot y \leq U \land 1 \leq x \land m_r \leq U \land x \cdot y \neq m_r
\]
where $L = N \idiv 2, U = 1 + L$, $m_r = (x \cdot (\mathit{ite}(y \geq 0, y, N + y))) \mod N$.
We assume $N$ is even, such as $N = 2^{256}$.
The solver associates a variable $m$ with $x \cdot y$ and $m'$ with $x \cdot y'$ and $y'$ with $\mathit{ite}(y \geq 0, y, N + y)$
and includes the constraints $0 \leq m_r < N, m_q \cdot N + m_r = m'$, where $m_q$ is an integer variable.
The most interesting case is where $y < 0$, so $y' = y + N$. Gr{\"o}bner basis completion
then allows to derive $m_q N + m_r = m' = x(y + N) = xy + xN = m + xN$, which by integer linear arithmetic reasoning (the extended GCD test)
contradicts $m \neq m_r$ because the absolute value of both variables is below $N$.
\end{example}

Our extraction of linear constraints represents a partial integration of linear programming and polynomial arithmetic,
that favors only including linear inequalities over variables and monomials that are already present. Our implementation
does not include any variables for new monomials produced by completion.
In comparison, the approach in~\cite{DBLP:journals/pacmpl/KincaidKZ23} proposes a domain for abstract interpretation that populates a linear
solver with all equations produced by a completion. We have not experimented in depth with extending our approach with a full basis,
or use it as a starting point for finding lemmas based on Positivstellensatz or other extension mechanisms~\cite{DBLP:conf/csl/Tiwari05,DBLP:conf/cade/PlatzerQR09}.





\usetikzlibrary{shapes,arrows}
\usetikzlibrary{positioning}

\tikzstyle{block} = [rectangle, draw, text centered, rounded corners, minimum height=2em]
\tikzstyle{line} = [draw, -latex']

\begin{wrapfigure}{r}{0.4\textwidth} 
  \begin{center}
    \begin{tikzpicture}[node distance = 3em, scale = 0.2]
  \node [circle, draw] (xroot) {$x$};
  \node [circle, below of = xroot, left of= xroot, draw] (xnext) {$x$};
  \node [circle, below of = xnext, left of= xnext, draw] (y1) {$y$};
  \node [circle, below of = xnext, right of = xnext] (yempty) {};
  \node [circle, right of= yempty, draw] (y2) {$y$};
  \node [circle, below of = y1, left of = y1] (t1) {$5$};
  \node [circle, below of = y1, right of = y1] (t2) {$0$};
  \node [circle, right of = t2] (t3) {$1$};
  \node [circle, below of = y2, right of = y2] (t4) {$1$};
  \path [line] (xroot.west) -- (xnext.north);
  \path [line] (xroot.east) -- (y2.north);
  \path [line] (xnext.west) -- (y1.north);
  \path [line] (xnext.east) -- (y2.north);
  \path [line] (y1.west) -- (t1.north);
  \path [line] (y1.east) -- (t2.north);
  \path [line] (y2.west) -- (t3.north);
  \path [line] (y2.east) -- (t4.north);

\end{tikzpicture}

  \end{center}
  \vspace{1pt}
  \caption{PDD representation of $5x^2y + xy + y + x + 1$\label{fig:pdd}}
\end{wrapfigure}
We use an adaptation of ZDD (Zero suppressed decision diagrams~\cite{Minato93,NishinoYMN16}) to represent polynomials.
The representation has the advantage that polynomials are stored in a shared data-structure and operations
over polynomials are memorized. A polynomial over the real is represented as an acyclic graph, where 
nodes are labeled by variables and edges are labeled by coefficients. Figure~\ref{fig:pdd} shows a polynomial stored
in a polynomial decision diagram, PDD.


The root node labeled by $x$ represents the polynomial $x\cdot l + r$, 
where $l$ is the polynomial of the left sub-graph and $r$ the polynomial
of the right sub-graph. The left sub-graph is allowed to be labeled again by $x$, 
but the right sub-graph may only have nodes labeled by variables that are smaller
in a fixed ordering. The fixed ordering used in this example sets $x$ above $y$.
Then the polynomial for the right sub-graph is $y + 1$, and the polynomial with the
left sub-graph is $5xy + (y + 1)$.

