\section{Non-Linear Arithmetic}
\label{sec:nla}

Similar to solving for integer feasibility, the arithmetic solver
solves constraints over polynomials using a waterfall model for non-linear
constraints.
At the basis it maintains, for every monomial 
term $x \cdot x \cdot y$, a definition
$m = x \cdot x \cdot y$, where $m$ is a variable
that represents the monomial $x \cdot x \cdot y$.
The module for non-linear arithmetic then collects the monomial
definitions that are violated by the current evaluation,
that is $\eval{m} \neq \eval{x} \cdot \eval{x} \cdot \eval{y}$. It  attempts to 
establish a valuation $\beta'$ where $\beta'(m) = \beta'(x) \cdot \beta'(x) \cdot \beta'(y)$, 
or derive a consequence that no such evaluation exists.



\subsection{Patch Monomials}

A \emph{patch} for a variable $x$ is \emph{admissible} if the update $\eval{x} := v$
does not break any integer linear constraints and $x$ does not occur in monomial
equations that are not already false under $\beta$.

\begin{itemize}
\item Set $\eval{m} := \eval{x} \cdot \eval{x} \cdot \eval{y}$ and check if the patch of $m$ is admissible.
\item Try to set $\eval{y} := \eval{m} / (\eval{x} \cdot \eval{x})$, provided $\beta{x}$ is not 0,
  and check that the patch for $x$ is admissible.
\item When $\eval{m} = r^2$ for a rational and $m := x \cdot x$ try patching $x$ by setting $\eval{x} := \pm r$.
\end{itemize}


\subsection{Bounds propagation}
A relatively inexpensive step is to propagate and check bounds based 
on non-linear constraints. For example, for $y \geq 3$, then $m = x\cdot x\cdot y \geq 3$,
if furthermore $x \leq -2$, we have the strengthened bound $m \geq 12$.
Bounds propagation can also flow from bounds on $m$ to bounds on the 
variables that make up the monomial, such that when $m \geq 8, 1 \leq y \leq 2, x \leq 0$, 
then we learn the stronger bound $x \leq -2$ on $x$. It uses an interval arithmetic abstraction,
that understands bounds propagation over squares. Thus, if $-2 \leq x \leq 2$, then $0 \leq x^2 \leq 4$
instead of $-4 \leq x^2 \leq 4$.

The solver also performs Horner expansions of polynomials to derive stronger bounds.
For example, if $x \geq 2, y \geq -1, z \geq 2$, then $y + z \geq 1$ 
and therefore $x\cdot (y + z) \geq 2$, but we would not be
able to deduce this fact if combining bounds individually for $x\cdot y$ 
and $x \cdot z$ because no bounds can be inferred for $x \cdot y$ in isolation.
The solver therefore attempts different re-distribution of multiplication
in an effort to find stronger bounds.


\subsection{Adding bounds}
Non-linear bounds propagation only triggers if all variables are either bounded from above or below or occur with an even power.
The solver includes a pass where it adds a bound $x \geq 0$ to variables $x$ where $lo_x = -\infty, hi_x = +\infty$.
The added bound may help trigger bounds propagation, such as detecting conflicts on $xy > 0, xz > 0, y > 0 > z$.


\input{grobner}


\subsection{Incremental linearization}
Following~\cite{CimattiGIRS18} we incrementally linearize monomial definitions that
currently evaluate to false.
For example, we include lemmas of the form $x = 0 \rightarrow m = 0$
and $x = 1 \rightarrow m = y$, for $m = x^2y$.
Incremental linearization proceeds by first applying linearizations that are considered cheap,
such as case splitting on whether variables take values 0, 1, -1, when these boundary conditions
are exhausted, instantiates lemmas based on monotonicity of multiplication and tangents.
It is possible that there are overlapping monomial definitions, such as $m' = x \cdot y$.
Then incremental linearization takes into account that the definition for $m$ can be
\emph{factored} into $m' \cdot x$. It also uses specialized congruence closure reasoning,
recognizing equalities modulo signs, such that when $m = x \cdot y, m' = z \cdot y$ and
$x = -z$ in the current context, then $m \sim -m'$.

% LN, not sure if it is a good place, and 
To find all factorizations of monomial $m = \prod_{i \in A}{x_i} $ as $m = m_0 \cdot m_1$, we choose $a \in A$ and enumerate over all proper subsets $B$ of $A$ containing $a$. For each $B$ we check that $m_0 = \prod_{i \in B}{x_i}$ and $m_1 = \prod_{i \in A \setminus B}{x_i}$ are monomials.

To support floating point arithmetic reasoning
we also include incremental linearization lemmas for special
cases of exponentiation~\cite{DBLP:journals/tocl/CimattiGIRS18}. 
We also added rules for incremental linearization of divisibility operations. 
The front-end to the core arithmetic solver axiomatizes integer and real division operations using multiplication and addition,
so that the solver does not have to reason about division. Nevertheless, we found use cases for
axioms of the form $y > 0 \land x > z \implies x/y > z/y$, bypassing indirect reasoning around constraints created by axioms.



\subsection{NLSat}
As an end-game attempt, the solver attempts to solver the non-linear constraints using a complete solver
for Tarski's fragment supported by the NLSat solver~\cite{JovanovicM12}. NLSAT is complete for non-linear arithmetic
and includes branch-and-bound to handle cases of integer arithmetic. It can therefore sometimes be used to solve
goals, bypassing the partial heuristics entirely. The solver therefore includes selected calls to NLSat with a small
resource bound to close branches before attempting incomplete heuristics such as incomplete linearization.
The results in Section~\ref{sec:eval} suggests that our use of NLSat with a resource bound currently incurs
significant overhead on easy problems, but overall is an advantage. We found that it is sometimes the case that turning off NLSat
all-together can speed up the solver significantly, but is overall a disadvantage.

